---
title: "Moneyball EDA: On-Base Skill, Run Creation, and Wins (Annotated)"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

## Ghid de citire

Acest fișier este partea „poveste + dovezi” a proiectului.

Aici demonstrăm ideea Moneyball:

- `OBP` și `SLG` ajută la producerea de run-uri;
- run-urile marcate și prevenite influențează victoriile;
- Oakland 2001-2002 este un exemplu interesant în acest context.

Ca să-l înțelegi ușor:

- urmărești mai întâi ce variabile sunt puse una contra alteia;
- apoi te uiți la linia de trend și la corelație;
- la final citești interpretarea și vezi ce concluzie susține graficul.

## Setup

În `Setup` citim datele și pregătim două lucruri:

- o coloană care marchează Oakland 2001-2002;
- o funcție ajutătoare pentru scatter plot-uri cu linie de regresie și corelație.

```{r}
# Save the data file path in a variable.
data_source <- "baseball.csv"

# Load the CSV into a data frame named `baseball`.
baseball <- read.csv(data_source, stringsAsFactors = FALSE)

# Convert categorical columns into factors.
baseball$Team <- as.factor(baseball$Team)
baseball$League <- as.factor(baseball$League)
baseball$Playoffs <- as.factor(baseball$Playoffs)

# Create a logical column that identifies Oakland in 2001 and 2002.
baseball$Oakland_0102 <- baseball$Team == "OAK" & baseball$Year %in% c(2001, 2002)
```

```{r}
# Define a helper function to draw a scatter plot, fit a linear trend,
# and optionally highlight selected observations.
plot_scatter_with_lm <- function(data, xvar, yvar,
                                 xlab = xvar,
                                 ylab = yvar,
                                 main = NULL,
                                 highlight = NULL,
                                 highlight_col = "#d7301f",
                                 base_col = "#6baed6") {
  # Keep only rows where both plotted variables are present.
  complete_idx <- complete.cases(data[, c(xvar, yvar)])
  plot_data <- data[complete_idx, ]

  # Extract the x and y vectors from the filtered data.
  x <- plot_data[[xvar]]
  y <- plot_data[[yvar]]

  # Fit a simple linear regression line.
  fit <- lm(y ~ x)

  # Compute the Pearson correlation between x and y.
  r_value <- cor(x, y)

  # Create default aesthetics for all points.
  point_col <- rep(base_col, nrow(plot_data))
  point_pch <- rep(16, nrow(plot_data))
  point_cex <- rep(0.8, nrow(plot_data))

  # If a highlight vector is provided, update those points' appearance.
  if (!is.null(highlight)) {
    highlight_vals <- highlight[complete_idx]
    point_col[highlight_vals] <- highlight_col
    point_pch[highlight_vals] <- 19
    point_cex[highlight_vals] <- 1.2
  }

  # Draw the scatter plot.
  plot(x, y,
       pch = point_pch,
       col = point_col,
       cex = point_cex,
       xlab = xlab,
       ylab = ylab,
       main = main)

  # Add the fitted regression line.
  abline(fit, col = "#08519c", lwd = 2)

  # Create the text label that shows the Pearson correlation.
  legend_text <- paste0("Pearson r = ", format(round(r_value, 3), nsmall = 3))

  # Read the current plot limits so the text can be placed inside the panel.
  usr <- par("usr")

  # Add the correlation label near the upper-left corner of the plot.
  text(x = usr[1] + 0.04 * (usr[2] - usr[1]),
       y = usr[4] - 0.06 * (usr[4] - usr[3]),
       labels = legend_text,
       adj = c(0, 1),
       cex = 0.9)

  # Return useful objects invisibly so they can be saved without printing.
  invisible(list(correlation = r_value, model = fit, data = plot_data))
}
```

## Which Offensive Metric Best Predicts Wins?

Aici comparăm trei statistici ofensive:

- `OBP`
- `SLG`
- `BA`

Scopul este să vedem care dintre ele este mai strâns legată de `W`.

The classic Moneyball question is whether some offensive measures are more informative about winning than others. The next set of plots compares `OBP`, `SLG`, and `BA` against total season wins.

```{r}
# Save current plotting settings.
old_par <- par(no.readonly = TRUE)

# Restore them automatically after plotting.
on.exit(par(old_par), add = TRUE)

# Create a 2-by-2 panel layout.
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Plot OBP against wins and highlight Oakland 2001-2002.
obp_w <- plot_scatter_with_lm(
  baseball, "OBP", "W",
  xlab = "On-Base Percentage (OBP)",
  ylab = "Wins (W)",
  main = "OBP vs W",
  highlight = baseball$Oakland_0102
)

# Plot SLG against wins and highlight Oakland 2001-2002.
slg_w <- plot_scatter_with_lm(
  baseball, "SLG", "W",
  xlab = "Slugging Percentage (SLG)",
  ylab = "Wins (W)",
  main = "SLG vs W",
  highlight = baseball$Oakland_0102
)

# Plot BA against wins.
ba_w <- plot_scatter_with_lm(
  baseball, "BA", "W",
  xlab = "Batting Average (BA)",
  ylab = "Wins (W)",
  main = "BA vs W"
)

# Create an empty panel for a legend.
plot.new()

# Add a legend that explains colors and symbols.
legend("center",
       legend = c("All teams", "Oakland 2001-2002", "Linear trend"),
       col = c("#6baed6", "#d7301f", "#08519c"),
       pch = c(16, 19, NA),
       lty = c(NA, NA, 1),
       lwd = c(NA, NA, 2),
       bty = "n")
```

Interpretation:

- `OBP` has the strongest correlation with wins among the three variables (Pearson r = 0.482), so it is the best single predictor of `W` in this comparison.
- `BA` (Pearson r = 0.409) and `SLG` (Pearson r = 0.402) are both positively related to wins, but their associations are weaker than `OBP`.
- This supports the original Moneyball idea that getting on base captures offensive value more effectively than batting average alone.
- Even so, none of these correlations is perfect, which means wins also depend on run prevention, sequencing, roster balance, and other season-level factors.

## Runs Scored and Runs Allowed as Direct Drivers of Wins

După ce comparăm statistici ofensive, trecem la mecanismul direct al victoriilor:

- dacă marchezi mai multe run-uri, ai șanse mai mari să câștigi;
- dacă primești mai puține run-uri, ai șanse mai mari să câștigi.

If offensive skill matters because it creates runs, then the most direct relationship should appear between wins and the run totals themselves.

```{r}
# Save current plotting settings.
old_par <- par(no.readonly = TRUE)

# Restore them automatically after plotting.
on.exit(par(old_par), add = TRUE)

# Create a one-row, two-column plot layout.
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Plot runs scored against wins.
rs_w <- plot_scatter_with_lm(
  baseball, "RS", "W",
  xlab = "Runs Scored (RS)",
  ylab = "Wins (W)",
  main = "RS vs W"
)

# Plot runs allowed against wins.
ra_w <- plot_scatter_with_lm(
  baseball, "RA", "W",
  xlab = "Runs Allowed (RA)",
  ylab = "Wins (W)",
  main = "RA vs W"
)
```

Interpretation:

- `RS` is positively associated with wins (Pearson r = 0.512): teams that score more runs usually win more games.
- `RA` is negatively associated with wins (Pearson r = -0.532): teams that allow fewer runs usually win more games.
- The slightly stronger absolute correlation for `RA` suggests run prevention is at least as important as run creation in explaining season win totals.
- This is the core accounting identity behind baseball success: wins emerge from the balance between scoring and preventing runs.

## The Moneyball Chain: `OBP` and `SLG` Lead to `RS`, Which Leads to `W`

Aceasta este secțiunea-cheie din punct de vedere conceptual.

Lanțul logic este:

- jucătorii ajung pe bază și lovesc puternic;
- asta produce `RS`;
- iar `RS` contribuie la `W`.

The central Moneyball claim is not only that offensive skills matter, but that they matter because they translate into run production.

```{r}
# Save current plotting settings.
old_par <- par(no.readonly = TRUE)

# Restore them when the plotting is done.
on.exit(par(old_par), add = TRUE)

# Create another two-panel layout.
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Plot OBP against runs scored.
obp_rs <- plot_scatter_with_lm(
  baseball, "OBP", "RS",
  xlab = "On-Base Percentage (OBP)",
  ylab = "Runs Scored (RS)",
  main = "OBP vs RS"
)

# Plot SLG against runs scored.
slg_rs <- plot_scatter_with_lm(
  baseball, "SLG", "RS",
  xlab = "Slugging Percentage (SLG)",
  ylab = "Runs Scored (RS)",
  main = "SLG vs RS"
)
```

Interpretation:

- `OBP` has an extremely strong positive relationship with run scoring (Pearson r = 0.901), which strongly supports the idea that getting on base is a direct engine of offensive production.
- `SLG` is also extremely strongly related to `RS` (Pearson r = 0.919), and in this dataset it is marginally stronger than `OBP` as a predictor of runs scored.
- Putting the pieces together: `OBP` and `SLG` help explain `RS`, and `RS` in turn helps explain `W`.
- The key Moneyball nuance is that `OBP` stands out more clearly when the target is wins, not just runs, because batting average misses walks and therefore misses part of the path from offensive events to victories.

## Oakland 2001-2002 in the Moneyball Plots

Aici facem legătura dintre analiză și povestea Moneyball.

Important:

- datasetul ne permite să vedem dacă Oakland era peste linia medie de performanță;
- dar nu ne permite să dovedim complet ideea de „per dollar”, pentru că nu avem payroll.

The red points mark the Oakland Athletics in 2001 and 2002 on the `OBP` vs `W` and `SLG` vs `W` plots.

Interpretation:

- Oakland 2001 and 2002 sit well above the fitted lines, meaning they won substantially more games than the average team with similar `OBP` or `SLG`.
- Relative to the overall `OBP`-wins relationship, Oakland overperformed by roughly 14 to 17 wins; relative to the `SLG`-wins relationship, they also overperformed by roughly 15 to 17 wins.
- That pattern is consistent with the idea that Oakland translated strong underlying skills into unusually high win totals.
- However, this dataset does **not** include payroll, so it cannot answer whether Oakland got more wins **per dollar** of `OBP` than average. It can only show that Oakland got more wins than typical teams with similar offensive rates.

## Conclusion

Concluzia acestui fișier este una foarte importantă pentru prezentare:

- `OBP` este mai bun decât `BA` ca predictor pentru `W`;
- `OBP` și `SLG` sunt strâns legate de `RS`;
- `RS` și `RA` sunt motorul direct al victoriilor.

The recreated Moneyball analysis supports the classic logic of the story. Among `OBP`, `SLG`, and `BA`, `OBP` is the strongest single predictor of wins. More fundamentally, `OBP` and `SLG` strongly predict runs scored, while both `RS` and `RA` directly shape team wins. Oakland's 2001-2002 seasons sit above the usual offensive-to-wins trend, which reinforces the idea that those teams converted their offensive profile into exceptional results, even if payroll efficiency cannot be verified from this dataset alone.
