---
title: "Playoff vs Non-Playoff Team Profiles (Annotated)"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

## Ghid de citire

Acest fișier compară profilul echipelor care ajung în playoff cu profilul celor care nu ajung.

Întrebarea centrală este:

- ce statistici separă cel mai bine echipele `Playoffs = 1` de echipele `Playoffs = 0`?

Te uiți la:

- densități pentru formă generală;
- boxplot-uri pentru mediană și dispersie;
- medii cu SD pentru comparație directă;
- `RS` vs `RA` pentru o imagine intuitivă a performanței.

## Setup

În `Setup` citim datele, transformăm variabilele categorice și pregătim forma „lungă” a datasetului pentru graficele faceted.

```{r}
# Load the packages used in this chapter.
library(dplyr)
library(tidyr)
library(ggplot2)

# Store the CSV path in a variable.
data_source <- "baseball.csv"

# Read the data and convert important categorical columns to factors.
baseball <- read.csv(data_source, stringsAsFactors = FALSE) |>
  mutate(
    League = factor(League),
    Playoffs = factor(Playoffs, levels = c(0, 1))
  )

# List the features we want to compare across playoff groups.
comparison_features <- c("W", "RS", "RA", "OBP", "SLG", "BA")

# Convert the data from wide format to long format for faceted plotting.
comparison_long <- baseball |>
  select(Playoffs, all_of(comparison_features)) |>
  pivot_longer(
    cols = all_of(comparison_features),
    names_to = "feature",
    values_to = "value"
  )
```

## Overlapping Density Plots by Playoff Status

Această secțiune îți arată forma completă a distribuțiilor pentru playoff vs non-playoff.

These density plots compare the full distributions of core performance statistics for playoff and non-playoff teams.

```{r}
# Create faceted density plots, coloring by playoff status.
ggplot(comparison_long, aes(x = value, colour = Playoffs, fill = Playoffs)) +
  geom_density(alpha = 0.2, linewidth = 0.9, na.rm = TRUE) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  scale_colour_manual(values = c("0" = "#6baed6", "1" = "#cb181d")) +
  scale_fill_manual(values = c("0" = "#6baed6", "1" = "#cb181d")) +
  labs(
    title = "Density of Key Statistics by Playoff Status",
    subtitle = "Playoffs = 1 indicates a playoff team",
    x = NULL,
    y = "Density"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- `W` shows the clearest separation: playoff teams are concentrated at much higher win totals, while non-playoff teams cluster lower.
- `RS` and `RA` also separate strongly, with playoff teams shifted toward more runs scored and fewer runs allowed.
- `OBP` and `SLG` show visible but more moderate separation, consistent with offensive quality helping teams qualify for the postseason.
- `BA` does differ between groups, but the overlap is larger, so batting average alone is a weaker separator than broader run-creation measures.

## Box Plots Split by Playoff Status

Aici accentul cade pe mediană, IQR și outlieri.

Box plots make it easier to compare medians, spread, and outliers across the same six statistics.

```{r}
# Draw box plots for each feature, split by playoff status.
ggplot(comparison_long, aes(x = Playoffs, y = value, fill = Playoffs)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.5, na.rm = TRUE) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("0" = "#9ecae1", "1" = "#fb6a4a")) +
  labs(
    title = "Box Plots of Key Statistics by Playoff Status",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

Interpretation:

- The median playoff team wins far more games than the median non-playoff team, and the interquartile ranges overlap only modestly.
- Playoff teams tend to have higher medians for `RS`, `OBP`, `SLG`, and `BA`, but lower medians for `RA`.
- The shift in `RA` is especially important because preventing runs is one of the most visible structural differences between playoff and non-playoff clubs.
- Outliers exist in both groups, but the median shifts matter more than a few unusual seasons.

## Mean Comparison with Standard Deviation Error Bars

Această secțiune este utilă când vrei să compari rapid nivelul mediu al fiecărei statistici între cele două grupuri.

The next chart compares group means and one-standard-deviation error bars for each statistic.

```{r}
# Summarize each feature by playoff status using mean and standard deviation.
comparison_summary <- comparison_long |>
  group_by(Playoffs, feature) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    .groups = "drop"
  )

# Draw grouped bars for the means and error bars for one standard deviation.
ggplot(comparison_summary, aes(x = feature, y = mean_value, fill = Playoffs)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_value - sd_value, ymax = mean_value + sd_value),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  scale_fill_manual(values = c("0" = "#6baed6", "1" = "#cb181d")) +
  labs(
    title = "Mean Statistics for Playoff and Non-Playoff Teams",
    subtitle = "Error bars show one standard deviation",
    x = "Statistic",
    y = "Mean value"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- Playoff teams average about 18 more wins, nearly 89 more runs scored, and roughly 61 fewer runs allowed than non-playoff teams.
- `OBP` and `SLG` are also clearly higher for playoff teams, reinforcing the idea that better offensive process stats translate into stronger outcomes.
- `BA` rises for playoff teams too, but by a smaller margin than `OBP` and `SLG`.
- This makes `RS`, `RA`, `W`, `OBP`, and `SLG` the most informative separators of postseason qualification in this dataset.

## Runs Scored vs Runs Allowed by Playoff Status

Acest grafic este foarte bun pentru prezentare, pentru că se înțelege vizual foarte repede:

- echipele bune tind să fie sus și în stânga;
- echipele slabe tind să fie jos și în dreapta.

This scatter plot places offense and defense on the same diagram. Teams in the upper-left region score many runs and allow few runs, so they should be the strongest playoff candidates.

```{r}
# Plot runs allowed on the x-axis and runs scored on the y-axis.
# Color the points by playoff status.
ggplot(baseball, aes(x = RA, y = RS, colour = Playoffs)) +
  geom_point(alpha = 0.75, size = 2) +
  scale_colour_manual(values = c("0" = "#6baed6", "1" = "#cb181d")) +
  labs(
    title = "Runs Scored vs Runs Allowed",
    subtitle = "Playoffs = 1 indicates a playoff team",
    x = "Runs Allowed (RA)",
    y = "Runs Scored (RS)"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- Playoff teams cluster more heavily in the upper-left region of the plot, where teams combine strong offense with strong run prevention.
- Non-playoff teams are much more common in the lower-right region, where they score fewer runs and allow more.
- This plot makes the core separation especially intuitive: postseason teams usually create a positive run differential profile.
- Overall, `W`, `RS`, and `RA` provide the clearest separation, while `OBP` and `SLG` remain the strongest underlying offensive indicators behind that gap.

## Conclusion

Ideea mare de reținut:

- echipele de playoff sunt separate mai clar prin `W`, `RS` și `RA`;
- `OBP` și `SLG` adaugă și ele diferențe importante;
- `BA` ajută, dar mai puțin.

Playoff teams differ from non-playoff teams most clearly on wins, runs scored, and runs allowed, with `OBP` and `SLG` also showing meaningful separation. Batting average moves in the same direction, but it is less decisive. The broad pattern is exactly what baseball logic suggests: playoff teams tend to score more, prevent more, and therefore accumulate far more wins.
