---
title: "Correlation Structure of Baseball Statistics (Annotated)"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

## Ghid de citire

Acest fișier arată relațiile dintre variabile sub formă de corelații.

Întrebările importante sunt:

- ce variabile „merg împreună”;
- ce variabile sunt cele mai legate de `W`;
- cum se văd rolurile aproape simetrice ale ofensivei și apărării.

## Setup

În `Setup` păstrăm doar variabilele numerice, pentru că doar între ele putem calcula corelații Pearson.

```{r}
# Load the packages used in this chapter.
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)

# Save the CSV path in a variable.
data_source <- "baseball.csv"

# Read the dataset into a data frame.
baseball <- read.csv(data_source, stringsAsFactors = FALSE)

# Keep only numeric columns for correlation analysis.
numeric_data <- baseball |>
  select(where(is.numeric))
```

## Correlation Matrix Heatmap Ordered by Hierarchical Clustering

Aceasta este imaginea de ansamblu a relațiilor dintre variabile.

The heatmap below shows pairwise Pearson correlations among all numeric variables. The variables are reordered using hierarchical clustering so related statistics appear near each other.

```{r}
# Compute the pairwise correlation matrix.
# `pairwise.complete.obs` uses all available paired observations.
corr_mat <- cor(numeric_data, use = "pairwise.complete.obs")

# Make a copy of the correlation matrix for clustering only.
# We do this because clustering cannot handle NA values.
corr_for_clustering <- corr_mat

# Replace any missing correlations with 0 only for clustering purposes.
corr_for_clustering[is.na(corr_for_clustering)] <- 0

# Force the diagonal to 1 because each variable is perfectly correlated with itself.
diag(corr_for_clustering) <- 1

# Perform hierarchical clustering on a distance matrix derived from correlations.
cluster_order <- hclust(as.dist(1 - corr_for_clustering))$order

# Get the variable names in clustered order.
ordered_names <- colnames(corr_mat)[cluster_order]

# Reorder the original correlation matrix using the clustered order.
corr_ordered <- corr_mat[ordered_names, ordered_names]

# Convert the matrix into long format so ggplot can draw a heatmap.
corr_long <- as.data.frame(as.table(corr_ordered)) |>
  rename(feature_x = Var1, feature_y = Var2, correlation = Freq)

# Draw the heatmap.
ggplot(corr_long, aes(x = feature_x, y = feature_y, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Correlation Heatmap of Numeric Features",
    x = NULL,
    y = NULL,
    fill = "Pearson r"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

Interpretation:

- Offensive variables cluster together, especially `OBP`, `SLG`, `BA`, and `RS`.
- Defensive variables such as `RA`, `OOBP`, and `OSLG` also group together and tend to show opposite-sign relationships with winning.
- `W` sits between offensive and defensive performance, reflecting the fact that wins are jointly driven by both scoring and prevention.
- The clustered structure makes the dataset's baseball logic very visible: related skills move together.

## Correlation of Each Feature with `W`

Aici izolăm întrebarea cea mai practică: **ce este cel mai legat de numărul de victorii?**

This chart ranks each numeric feature by its Pearson correlation with wins.

```{r}
# Compute the correlation between every numeric feature and W.
cor_with_w <- numeric_data |>
  summarise(across(everything(), ~ cor(.x, baseball$W, use = "pairwise.complete.obs"))) |>
  pivot_longer(
    cols = everything(),
    names_to = "feature",
    values_to = "correlation"
  ) |>
  filter(feature != "W") |>
  arrange(desc(correlation))

# Draw a horizontal bar chart sorted by correlation value.
ggplot(cor_with_w, aes(x = fct_reorder(feature, correlation), y = correlation, fill = correlation)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) +
  labs(
    title = "Correlation of Each Numeric Feature with Wins",
    x = NULL,
    y = "Pearson r"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

Interpretation:

- `RS`, `OBP`, `BA`, and `SLG` all correlate positively with wins, but `RS` is the strongest positive relationship among the core offensive statistics.
- `RA`, `OOBP`, and `OSLG` show strong negative correlations, highlighting the importance of run prevention.
- The strongest single relationship in absolute terms is with opponent on-base prevention (`OOBP`) among the variables available for fewer seasons, while `RA` remains the most direct full-sample defensive summary.
- Overall, the ranking confirms that both offense and defense matter, with run-based measures closest to actual team success.

## `RS` vs `W` and `RA` vs `W` Side by Side

Această comparație e foarte bună pentru înțeles ideea de bază a baseball-ului:

- nu e suficient doar să marchezi;
- trebuie și să previi run-uri.

The next panel compares the near-symmetric roles of scoring runs and preventing them.

```{r}
# Reshape RS and RA into long format so they can be faceted side by side.
run_win_long <- baseball |>
  select(W, RS, RA) |>
  pivot_longer(
    cols = c(RS, RA),
    names_to = "feature",
    values_to = "value"
  )

# Plot each relationship in its own panel.
ggplot(run_win_long, aes(x = value, y = W)) +
  geom_point(alpha = 0.7, colour = "#636363") +
  geom_smooth(method = "lm", se = FALSE, colour = "#08519c", linewidth = 0.9) +
  facet_wrap(~ feature, scales = "free_x") +
  labs(
    title = "Runs Scored and Runs Allowed as Predictors of Wins",
    x = NULL,
    y = "Wins (W)"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- `RS` has a strong positive slope: teams that score more runs usually win more games.
- `RA` has a similarly strong negative slope: teams that allow more runs usually win fewer games.
- The two panels are close to mirror images, which underlines the nearly symmetric roles of offense and defense in producing wins.
- This symmetry is one of the clearest high-level insights in the dataset.

## `OBP` vs `SLG` Coloured by Win Tercile

Aici combinăm două statistici ofensive importante și vedem dacă echipele cu multe victorii tind să stea în zona de sus-dreapta.

This plot combines two core offensive rate statistics and colours teams by low-, middle-, and high-win terciles.

```{r}
# Divide teams into three equally sized groups based on wins.
baseball_terciles <- baseball |>
  mutate(
    win_tercile = ntile(W, 3),
    win_tercile = factor(
      win_tercile,
      levels = c(1, 2, 3),
      labels = c("Low wins", "Middle wins", "High wins")
    )
  )

# Plot OBP against SLG and color points by win tercile.
ggplot(baseball_terciles, aes(x = OBP, y = SLG, colour = win_tercile)) +
  geom_point(alpha = 0.75, size = 2) +
  scale_colour_manual(values = c("Low wins" = "#3182bd", "Middle wins" = "#9ecae1", "High wins" = "#cb181d")) +
  labs(
    title = "OBP and SLG by Win Tercile",
    x = "On-Base Percentage (OBP)",
    y = "Slugging Percentage (SLG)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- High-win teams are concentrated toward the upper-right of the plot, where both `OBP` and `SLG` are strong.
- Low-win teams cluster more often in the lower-left, reflecting weaker offensive profiles on both dimensions.
- There is still overlap across terciles, which shows that offense alone does not fully determine wins.
- Even so, teams that combine high on-base skill with high slugging do tend to win more consistently.

## Conclusion

Concluzia de reținut:

- variabilele ofensive se grupează între ele;
- variabilele defensive se grupează între ele;
- `W` este influențată de ambele blocuri de variabile.

The correlation analysis shows a clean and intuitive structure. Offense-related variables cluster together, defense-related variables cluster together, and wins depend on both sides of that ledger. Run scoring and run prevention are the most direct drivers of success, while `OBP` and `SLG` remain the strongest underlying rate-based offensive signals.
