---
title: "Tree-Based Regression Models for Predicting Wins (Annotated)"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

## Ghid de citire

Acesta este cel mai tehnic fișier din proiect.

Ca să-l înțelegi mai ușor, gândește-l în 6 pași:

1. citim datele și pachetele;
2. pregătim datele pentru model;
3. facem cross-validation;
4. căutăm cei mai buni parametri pentru Random Forest și XGBoost;
5. evaluăm modelele pe test;
6. vedem ce variabile au fost importante.

Regula bună aici este să nu citești doar codul, ci să te întrebi permanent:

- care este inputul?
- care este ținta?
- de ce facem acel pas înainte de modelare?

## Setup

În `Setup` ne asigurăm că:

- avem pachetele necesare;
- rezultatele pot fi reproduse;
- datele sunt citite corect pentru modelare.

```{r}
# List all packages required for this chapter.
required_packages <- c("tidymodels", "dplyr", "ggplot2", "vip", "ranger", "xgboost")

# Check which of those packages are missing on the current machine.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# If any packages are missing, install them from CRAN.
if (length(missing_packages) > 0) {
  install.packages(missing_packages, repos = "https://cloud.r-project.org")
}

# Load the modelling and plotting packages.
library(tidymodels)
library(dplyr)
library(ggplot2)
library(vip)

# Tell tidymodels to prefer its own versions of overlapping functions.
tidymodels_prefer()

# Set a random seed so results are reproducible.
set.seed(2026)

# Store the CSV path in a variable.
data_source <- "baseball.csv"

# Read the data and convert categorical variables to factors.
baseball <- read.csv(data_source, stringsAsFactors = FALSE) |>
  mutate(
    League = factor(League),
    Playoffs = factor(Playoffs)
  )
```

## Data Preparation

Aceasta este una dintre cele mai importante secțiuni.

Aici decidem:

- ce coloane intră în model;
- ce coloane se elimină;
- cum tratăm valorile lipsă;
- cum transformăm predictorii.

We predict season wins (`W`) using season-level performance statistics while removing identifier and post-hoc columns that should not be used as predictors.

```{r}
# Split the data into training and testing sets.
# 75% of rows go to training and 25% to testing.
win_split <- initial_split(baseball, prop = 0.75)
train_data <- training(win_split)
test_data <- testing(win_split)

# Define a preprocessing recipe for predicting W.
win_recipe <- recipe(W ~ ., data = train_data) |>
  # Remove identifier and leakage-prone variables.
  step_rm(Team, RankSeason, RankPlayoffs, Playoffs) |>
  # Convert League into dummy/indicator columns.
  step_dummy(League) |>
  # Fill missing OOBP values using the training median.
  step_impute_median(OOBP) |>
  # Fill missing OSLG values using the training median.
  step_impute_median(OSLG) |>
  # Remove predictors with zero variance.
  step_zv(all_predictors()) |>
  # Standardize numeric predictors to mean 0 and sd 1.
  step_normalize(all_numeric_predictors())

# Print the recipe summary.
win_recipe
```

Notes:

- `Team` is an identifier rather than a generalizable performance feature.
- `RankSeason`, `RankPlayoffs`, and `Playoffs` are post-hoc or target-adjacent variables and should not be used to predict wins.
- `League` is encoded with `step_dummy()`.
- `OOBP` and `OSLG` are imputed using median values estimated from the training data only.

## Cross-Validation

Cross-validation este pasul care ne ajută să alegem parametrii fără să folosim setul final de test.

We use 5-fold cross-validation on the training set to tune the models.

```{r}
# Reset the seed before making folds.
set.seed(2026)

# Create 5-fold cross-validation resamples from the training set.
cv_folds <- vfold_cv(train_data, v = 5)

# Print the fold object.
cv_folds
```

```{r}
# Define the performance metrics we want to track.
reg_metrics <- metric_set(rmse, mae, rsq)
```

## Random Forest Tuning

Aici definim modelul Random Forest și îl reglăm.

Parametrii principali sunt:

- `mtry` = câți predictori încearcă la fiecare split;
- `min_n` = numărul minim de observații într-un nod.

The random forest model tunes `mtry` and `min_n` over a regular 3 x 3 grid.

```{r}
# Define a random forest regression model with tunable parameters.
rf_spec <- rand_forest(
  mode = "regression",
  trees = 1000,
  mtry = tune(),
  min_n = tune()
) |>
  # Use the ranger engine and request permutation importance.
  set_engine("ranger", importance = "permutation")

# Combine the recipe and model into one workflow.
rf_wf <- workflow() |>
  add_recipe(win_recipe) |>
  add_model(rf_spec)

# Extract the tunable parameters and finalize them using the training data.
rf_params <- extract_parameter_set_dials(rf_wf) |>
  finalize(train_data)

# Build a regular grid with 3 levels for each tuning parameter.
rf_grid <- grid_regular(rf_params, levels = 3)

# Print the grid.
rf_grid
```

```{r}
# Reset the seed for reproducible tuning.
set.seed(2026)

# Tune the random forest over the cross-validation folds.
rf_tuned <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = reg_metrics
)

# Select the best parameter combination according to RMSE.
rf_best <- select_best(rf_tuned, metric = "rmse")

# Print the best tuning result.
rf_best
```

## XGBoost Tuning

Aici facem același lucru pentru XGBoost, dar modelul este mai flexibil și are mai mulți parametri de reglat.

The XGBoost model tunes `learn_rate`, `tree_depth`, and `mtry` with a random grid of 20 combinations.

```{r}
# Define a boosted tree regression model with tunable parameters.
xgb_spec <- boost_tree(
  mode = "regression",
  trees = 1000,
  learn_rate = tune(),
  tree_depth = tune(),
  mtry = tune(),
  min_n = 2,
  loss_reduction = 0,
  sample_size = 1
) |>
  # Use the xgboost engine.
  set_engine("xgboost")

# Create the workflow for the boosted tree model.
xgb_wf <- workflow() |>
  add_recipe(win_recipe) |>
  add_model(xgb_spec)

# Extract and finalize the tunable parameter set.
xgb_params <- extract_parameter_set_dials(xgb_wf) |>
  finalize(train_data)

# Reset the seed before generating the random tuning grid.
set.seed(2026)

# Create 20 random parameter combinations.
xgb_grid <- grid_random(xgb_params, size = 20)

# Print the grid.
xgb_grid
```

```{r}
# Reset the seed for reproducible tuning.
set.seed(2026)

# Tune XGBoost across the cross-validation folds.
xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid = xgb_grid,
  metrics = reg_metrics
)

# Select the best parameter set according to RMSE.
xgb_best <- select_best(xgb_tuned, metric = "rmse")

# Print the best parameters.
xgb_best
```

## Final Evaluation on the Test Set

Acum folosim pentru prima dată setul de test.

Foarte important:

- tuningul se face pe train + CV;
- evaluarea finală se face pe test.

After tuning, each workflow is finalized on the training data and evaluated on the held-out test set.

```{r}
# Insert the best RF parameters into the workflow.
final_rf_wf <- finalize_workflow(rf_wf, rf_best)

# Insert the best XGBoost parameters into the workflow.
final_xgb_wf <- finalize_workflow(xgb_wf, xgb_best)

# Fit the final RF workflow on the training set and evaluate on the test set.
rf_last_fit <- last_fit(final_rf_wf, split = win_split, metrics = reg_metrics)

# Fit the final XGBoost workflow on the training set and evaluate on the test set.
xgb_last_fit <- last_fit(final_xgb_wf, split = win_split, metrics = reg_metrics)

# Collect the evaluation metrics from both models into one table.
evaluation_table <- bind_rows(
  collect_metrics(rf_last_fit) |> mutate(model = "Random Forest"),
  collect_metrics(xgb_last_fit) |> mutate(model = "XGBoost")
) |>
  select(model, .metric, .estimate) |>
  arrange(.metric, model)

# Print the evaluation table.
evaluation_table
```

```{r}
# Collect predictions from both final models and combine them.
prediction_results <- bind_rows(
  collect_predictions(rf_last_fit) |> mutate(model = "Random Forest"),
  collect_predictions(xgb_last_fit) |> mutate(model = "XGBoost")
)

# Plot actual wins against predicted wins.
ggplot(prediction_results, aes(x = W, y = .pred)) +
  geom_point(alpha = 0.75, colour = "#3182bd") +
  geom_abline(intercept = 0, slope = 1, colour = "#cb181d", linetype = "dashed") +
  facet_wrap(~ model) +
  labs(
    title = "Predicted vs Actual Wins on the Test Set",
    x = "Actual wins",
    y = "Predicted wins"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- RMSE is the primary tuning metric, while MAE and R^2 provide complementary views of predictive quality.
- A strong model should produce predictions clustered near the 45-degree reference line in the scatter plot.
- Any visible bias at the extremes would indicate difficulty predicting very weak or very strong teams.

## Variable Importance

După ce vedem cât de bun e modelul, vrem să înțelegem și **de ce** prezice bine.

Aici apare ideea de importanță a variabilelor:

- ce predictori au influențat cel mai mult modelul;
- dacă domină `RS` și `RA`;
- sau dacă și `OBP` / `SLG` rămân importante.

We now inspect which predictors each tuned model relies on most heavily.

```{r}
# Fit the finalized RF workflow on the full training data.
rf_fit_train <- fit(final_rf_wf, data = train_data)

# Fit the finalized XGBoost workflow on the full training data.
xgb_fit_train <- fit(final_xgb_wf, data = train_data)

# Compute variable importance for the RF model.
rf_vi <- vip::vi(extract_fit_parsnip(rf_fit_train)$fit) |>
  mutate(model = "Random Forest")

# Compute variable importance for the XGBoost model.
xgb_vi <- vip::vi(extract_fit_parsnip(xgb_fit_train)$fit) |>
  mutate(model = "XGBoost")

# Combine both importance tables and keep the top 10 variables for each model.
combined_vi <- bind_rows(rf_vi, xgb_vi) |>
  group_by(model) |>
  slice_max(order_by = Importance, n = 10, with_ties = FALSE) |>
  ungroup()

# Print the combined importance table.
combined_vi
```

```{r}
# Reorder variables within each model for clearer plotting.
combined_vi_plot <- combined_vi |>
  arrange(model, desc(Importance)) |>
  mutate(
    variable_model = paste(model, Variable, sep = " | "),
    variable_model = factor(variable_model, levels = rev(unique(variable_model)))
  )

# Draw the importance plot for both models.
ggplot(combined_vi_plot, aes(x = variable_model, y = Importance, fill = model)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ model, scales = "free_y") +
  scale_x_discrete(labels = function(x) sub("^.* \\| ", "", x)) +
  labs(
    title = "Top 10 Variable Importance Features by Model",
    x = NULL,
    y = "Importance"
  ) +
  theme_minimal(base_size = 12)
```

Interpretation:

- In most reasonable tree-based models for this dataset, `RS` and `RA` should rank very highly because they are the most direct summaries of performance that drives wins.
- `OBP` and `SLG` may still appear among the top features, especially because they help explain offensive quality and run creation.
- If opponent statistics such as `OOBP` and `OSLG` rank highly, that would reinforce the importance of defensive prevention in season success.
- Comparing the two importance plots helps reveal whether the models agree on the same baseball story or emphasize different pathways.

## Conclusion

Concluzia de reținut din acest fișier:

- am construit două modele de regresie pentru `W`;
- le-am reglat corect prin cross-validation;
- le-am evaluat pe un set separat;
- apoi am interpretat ce variabile contează cel mai mult.

This modelling workflow uses a clean train/test split, consistent preprocessing, and cross-validated tuning to compare two tree-based regression models for wins. The final evaluation shows how accurately each model predicts held-out team seasons, while the importance analysis helps connect model performance back to baseball logic. In general, we expect direct run measures such as `RS` and `RA` to dominate, with offensive process metrics like `OBP` and `SLG` still contributing meaningful predictive signal.
